Normalization by Evaluation for Call-by-Push-Value and Focused Lambda-Calculus

### 2019-02-04 fscd19/fscd19.tex
[ fscd19 ] normal forms

In this section, we review the normalization by evaluation (NbE)
argument for the simply-typed lambda calculus with weak sums, setting
the stage for the later sections.
%
We work in a type-theoretic meta-language, with the basic judgement
$t : T$ meaning that object $t$ is an inhabitant of type $T$.
Meta-theoretic types can be thought of as sets, thus, the colon $:$
replaces the standard notion of elementhood $\in$, and we are free to
reuse $\in$ for other purposes.

@@

The category $\Cxt$ allows us to consider $A{\in}\_$ as a presheaf over
$\Cxt\op$ for any $A$, witnessed by \fbox{$\treindex : \Gamma \sublist
  \Delta \to A \in \Gamma \to A \in \Delta$}, which is the morphism
part of functor from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.

Simple types shall be given by the grammar
\fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \bone \mid S + T
  \mid \bzero$}.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the type $A \dashv \Gamma$,
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \dashv \Gamma}
\qquad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\qquad
  \nru{\tapp}
    {A \To B \from \Gamma \qquad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\bone \from \Gamma}
\qquad
  \nru{\tpair}
    {A_1 \from \Gamma \qquad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\qquad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\qquad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \qquad C \from \cext \Gamma A_1
      \qquad C \from \cext \Gamma A_2
    }{C \from \Gamma}
\qquad
  \nru{\tabort}
    {\bzero \from \Gamma}
    {C \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism.  Terms of type $A$ form a presheaf
$A{\from}\_$ as witnessed by the standard weakening operation%
\footnote{Here, $\tren$ is short for \emph{renaming}, but in a nameless
calculus we should better speak of \emph{reindexing}, which could, a
bit clumsily, also abbreviated to $\tren$.}
\fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
defined by recursion over $t : A \from \Gamma$.

\section{Call-By-Push Value}
\label{sec:cbpv}

>>>>

In this section, we review the normalization by evaluation (NbE)
argument for the simply-typed lambda calculus (STLC) with weak sums, setting
the stage for the later sections.
%
We work in a constructive type-theoretic meta-language, with the basic judgement
$t : T$ meaning that object $t$ is an inhabitant of type $T$.
Meta-theoretic types can be thought of as sets, thus, the colon $:$
replaces the standard notion of elementhood $\in$, and we are free to
reuse $\in$ for other purposes.

@@

The category $\Cxt$ allows us to consider $A{\in}\_$ as a presheaf over
$\Cxt\op$ for any $A$, witnessed by \fbox{$\treindex : \Gamma \sublist
  \Delta \to A \in \Gamma \to A \in \Delta$}, which is the morphism
part of functor from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
The associated naturality laws \fbox{$\treindex\,\tid\,x \equiv x$}
and \fbox{$\treindex\,\tau_2\,(\treindex\,(\tau_1\,x)) \equiv
\treindex\,(\tau_1 \comp \tau_2)\,x$} hold propositionally.

Simple types shall be given by the grammar
\fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \bone \mid S + T
  \mid \bzero$}.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the type \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\qquad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\qquad
  \nru{\tapp}
    {A \To B \from \Gamma \qquad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\bone \from \Gamma}
\qquad
  \nru{\tpair}
    {A_1 \from \Gamma \qquad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\qquad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\qquad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \qquad C \from \cext \Gamma A_1
      \qquad C \from \cext \Gamma A_2
    }{C \from \Gamma}
\qquad
  \nru{\tabort}
    {\bzero \from \Gamma}
    {C \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism.  A term such as
\(
  \tabs\,(\tabs\,(\tpair\,
    (\tvar\,(\tsuc\,\tzero))\,
    (\tabs\,(\tapp\,(\tvar\,(\tsuc\,\tzero))\,(\tvar\,\tzero)))))
\)
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to imagination of the reader,
as we shall not be concerned with considering
concrete terms in this article.

Terms of type $A$ form a presheaf
$A{\from}\_$ as witnessed by the standard weakening operation%
\footnote{Here, $\tren$ is short for \emph{renaming}, but in a nameless
calculus we should better speak of \emph{reindexing}, which could, a
bit clumsily, also abbreviated to $\tren$.}
\fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
defined by recursion over $t : A \from \Gamma$,
and naturality laws for $\tren$ analogously to $\treindex$.

Normal forms%
\footnote{
There is also a stronger
notion of normal form, requiring that two \emph{extensionally equal}
lambda-terms, \ie, those that denote the
same set-theoretical function, have the same normal form
\cite{mitchell:foundations,altenkirchDybjerHofmannScott:lics01,scherer:popl17}.
Such normal forms do not have a simple inductive definition, and we
shall not consider them in this article.
}
are logically characterized as those fulfilling the
\emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
Normal forms \fbox{$t : \Nf A \Gamma$} are mutually defined with
\emph{neutral} normal forms \fbox{$t : \Ne A \Gamma$}.
In the following inductive definition, we reuse the rule names from the
term constructors.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{\Ne A \Gamma}
\qquad
  \nru{\tabs}{\NfC B {\cext \Gamma A}}{\NfP {A \To B} \Gamma}
\qquad
  \nru{\tapp}
    {\NeP {A \To B} \Gamma \qquad \Nf A \Gamma}
    {\Ne B \Gamma}
\qquad
  \nru{\tne}{\Ne A \Gamma}{\Nf A \Gamma}
\\[2ex]
  \nru{\tunit}{}{\Nf \bone \Gamma}
\qquad
  \nru{\tpair}
    {\Nf {A_1} \Gamma \qquad \Nf {A_2} \Gamma}
    {\NfP {A_1 \times A_2} \Gamma}
\qquad
  \nru{\tprj_i}
    {\NeP {A_1 \times A_2} \Gamma}
    {\Ne {A_i} \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {\Nf {A_i} \Gamma}
    {\NfP {A_1 + A_2} \Gamma}
\quad
  \nru{\tcase}
    {\NeP {A_1 + A_2} \Gamma
      \quad \NfC C {\cext \Gamma A_1}
      \quad \NfC C {\cext \Gamma A_2}
    }{\Nf C \Gamma}
\quad
  \nru{\tabort}
    {\Ne \bzero \Gamma}
    {\Nf C \Gamma}
\end{gather*}
The new rule $\tne$ embeds $\tNe$ into $\tNf$,
\ie, allows us to forget neutrality.  Needless to say, $\tNe\,A$ and
$\tNf\,A$ are presheaves, \ie, support reindexing with $\tren$ just as terms do.
From a normal form we can extract the term via a overloaded function
\fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
\fbox{$\tm\_ : \Ne A \Gamma \to A \from \Gamma$}
that discards constructor $\tne$ but keeps all other constructors.
This erasure function naturally commutes with reindexing, making it a
natural transformation between the presheaves $\tNf\,A$ ($\tNe\,A$,
resp.) and $A{\from}\_$.  We shall simply write, for instance,
\fbox{$\tNf\,A \todot A{\from}\_$} for such presheaf morphisms.

\emph{Proof normalization} is concerned with finding a normal form
$u : \Nf A \Gamma$ for each derivation $t : A \from \Gamma$.  However,
\emph{term normalization} additionally requires that $\tm u \cong t$
with respect to a equational theory $\cong$ on terms.  Defining
\fbox{$\NF t = \exists u \of \tNf A\,\Gamma.\; \tm u \cong t$}
% for $t : A \from \Gamma$,
we aim to construct a normalization function
\[
  \tnorm : (t : A \from \Gamma) \to \NF t
\]
that gives us for each term $t$ a $\cong$-equal normal form $u$.
%\footnote{
A more accurate presentation of $\tNF$ would have been
$\sum_{u : \tNf A\,\Gamma} (\tm u \cong t)$ to make absolutely clear that we
want get hold of the witness $u$ of the existential statement;
however, we shall only use constructive
existentials\footnote{Constructive existence in
  contrast to classical existence and
  \emph{mere existence} from Homotopy Type Theory which
  requires propositional truncation.}
in this article and thus stick to the notation $\exists$.
%}
\section{Call-By-Push Value}
\label{sec:cbpv}
